\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-05-20 main.tex
read over - finalise figures - clean tex

\begin{abstract}

\end{abstract}

@@

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others,
increasing their payoff, at the expense of decreasing one’s own? In evolutionary
game theory, individuals are not required to be rational, instead they adapt
strategies based on mutation and exploration. Strategies are more likely to
spread if they have a high fitness either because the individuals who adopt them
have more offspring, or because they are imitated more often~\cite{Wu2015}. The
fitness of a strategy is not constant but depends on the composition of the
population. Individuals interact based on their strategies with other members of
the population and the payoffs they yield are translated into fitness.

It is commonly assumed that individuals compute their fitness after interacting
with a representative sample of the population, and remembering all the
interactions they participated in~\cite{Nowak2006}. Thus, they imply that
individuals have a perfect memory. However, when modelling how individuals make
decisions in each turn they are assumed to have very limited memory. To be
precise, most of the works in the literature focus on naive subjects who can
only choose from a restricted set of strategies~\cite{Nowak1992tit}, or who do
not remember anything beyond the outcome of the very last round~\cite{Baek2016}.
Note that there are a few notable exceptions~\cite{Hauert1997, Stewart2016}.
Thus, the perfect memory assumption is not only unrealistic but it also creates
this curious inconsistency.

This has led us to question the robustness of our understanding of cooperation,
In this work we explore whether cooperation can evolve if individuals compute
their fitness based on a minimum of social information. Though we are not the
first to question the assumptions of estimating fitness~\cite{Roca2006}, we are
the first to explore the effect of payoff memory.

@@

The remainder of the paper is organized as follows. In
section~\ref{section:model} we describe the model. In
section~\ref{section:results} we presents the results of the simulations.
Finally, section~\ref{section:conclusions} we outline the main conclusions.

@@

We consider a population of \(N\) players~\footnote{The terms ``player'' and
``individual'' are used interchangeably here.} (\(N\) is even) where mutations are
sufficiently rare. Thus, at any point in time there are at most two different
strategies present in the population; a \textit{resident} strategy and a
\textit{mutant} strategy. We assume a pairwise process where strategies spread
because they are imitated more often.

Each step of the evolutionary process consists of two stages, a game stage and
an updating stage. In the game stage each individual is randomly matched with
some other individual in the population to interact for a number of turns, where
subsequent turns occur with a fixed probability $\delta$. At each turn they
can choose to either cooperate (\(C\)) or to defect (\(D\)), and thus, at
each turn the possible outcomes are \(CC, CD, DC\) and \(DD\). The payoffs
depend on the outcome. If both cooperate they receive the reward payoff
\(R\), whereas if both defect they receive the punishment payoff \(P\).
If one cooperates but the other defects, the defector receives the temptation to
defect, \(T\), whereas the cooperator receives the sucker's payoff, \(S\).
We denote the payoffs of an individual as \(\mathcal{U} = (R, S, T, P)\).

We assume herein that individuals use \textit{reactive strategies} to make
decisions in each turn. Reactive strategies are a set of memory-one strategies
that only take into account the previous action of the opponent. They can be
written explicitly as a vector \(\in \R_{3}\), more specifically, a reactive
strategy \(s\) is given by \(s=(y, p, q)\) where \(y\) is the probability that
the strategy opens with a cooperation and \(p, q\) are the probabilities that
the strategy cooperates given that the opponent cooperated and defected
equivalently.

In the updating stage, two players are randomly drawn from the population, a
`learner' and a `role model'. Given that the learner's payoff $u_L\!\in\!
\mathcal{U}$ and that the role model's payoff $u_{RL}\!\in\! \mathcal{U}$, we
assume the learner adopts the role model's strategy based on the Fermi
distribution function,

@@

where $\beta\!\ge\!0$ is the relative influence of the payoffs on adopting the 
strategy of the other. We refer to $\beta$ as the intensity of selection.

This basic evolutionary step is repeated until either the mutant strategy goes
extinct, or until it fixes in the population. If the mutant fixes in the
population then the mutant strategy becomes the new resident strategy. After
either outcome we introduce a new mutant strategy uniformly chosen from all
reactive strategies at random, and we set the number of mutants to $1$. This
process of mutation and fixation/extinction is then iterated many times.

The perfect memory assumption occurs at the updating stage. The learner and the
role model are assumed to interact with a representative sample of the
population, and they remember all interactions they participate in. Thus, their
updating payoffs are based on the mean payoff they achieved over all the
interactions. These payoffs are referred to as the expected payoffs. We will
compare the expected payoffs to payoffs that are calculated when the role model
and learner do not remember all of their interactions. In order to account for
the effect of these different methods, we explore the cooperation rate within
the resident population over multiple generations. More details on our
methodology are found in Appendix~\ref{appendix:methods}.

@@

In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population. We compare
this to the classical scenario of the expected payoffs. We assume that each pair
of players interacts in a donation game. The donation game is a special case of
the prisoner's dilemma. Each player can choose to cooperate by providing a
benefit \(b\) to the other player at their cost \(c\), with \(0 < c < b\).
Thus, the payoffs of a one round interact are \(T=b, R=b-c, S=-c, P=0\).

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulations results
for the described process of section~\ref{section:model}.
Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving
conditional cooperation probabilities $p$ and $q$. The discount factor~$\delta$
is comparably high, thus the opening move \(y\) is a transient
effect and has no effect on the outcome. The left panel corresponds to the
standard scenario considered in the literature. It considers players who use
expected payoffs to update their strategies. The right panel shows the scenario
considered herein, in which players update their strategies based on their last
round’s payoff. The top panels assume a benefit \(b\) of 3, whereas the bottom
assume a benefit of 10.

The figure suggests that when updating is based on expected payoffs, players
tend to be more generous and more cooperative. The $q$-values of the resident
strategies are on average higher in the case of the expected payoffs. The
players will occasionally forgive a defection more often if their fitness
depends on interacting with every member of the population. On the other hand,
when social interactions are limited they are less forgiving. The average
cooperation rate for each simulation is calculated as the the average
cooperation rate within the resident population. In the case of the expected
payoffs, regardless the value of benefit, the average cooperation rate is
strictly higher than that of the last round payoffs. The difference based the
two methods is statistically significant, and the dissimilarity is even more
obvious for a high value of \(b\). More specifically, the average cooperation of
resident strategies drops from 97\% to 57\%.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.75\textwidth]{static/expected_and_stochastic_for_donation_game.pdf}\vspace{-3cm}
    \includegraphics[width=.75\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and last round with one interaction payoffs.} 
    We have run two simulations of the evolutionary process described in
    section~\ref{section:model} for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since simulations
    are run for a relatively high continuation probability of $\delta\!=\!0.999$, we
    do not report the players' initial cooperation probability $y$. The graphs show
    how often the resident population chooses each combination ($p,q$) of
    conditional cooperation probabilities in the subsequent rounds. ({\bf A}) If
    players update based on their expected payoffs, the resident population
    typically applies a strategy for which $p\!\approx\!1$ and
    $q\!\le\!1\!-\!c/b\!=\!0.9$. ({\bf B})
    When players update their strategies based on their realized payoffs in the last
    round, there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$ consistent
    with stable cooperation is somewhat smaller compared to the expected-payoff
    setting, $q\!<\!0.5$. The cooperation rate within the resident population
    (averaged over all games and over all time steps) is close to 100\%. Parameters:
    $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

We further explore the effect of the benefit.
Figure~\ref{fig:cooperation_rate_over_benefit} suggests that expected payoffs
always yield a higher cooperation rate. We observe that the cooperation rate
increases as the value of the benefit gets higher. In comparison, for the
limited memory payoffs, the cooperation rate remains unchanged, at approximately
50\%.

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_beta.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_beta.pdf}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
  \end{subfigure}
  \caption{{\bf The evolution of cooperation for different benefit values.} 
  We vary the benefit of defection $b$. In all cases, expected payoffs appear to
  overestimate the average cooperation rate the population achieves. ({\bf A})
  the probabilities \(p, q\) for resident population over \(10^7\) time steps
  for each benefit value. ({\bf B}) The cooperation rate within the resident population
  (averaged over all games and over all time steps) over the benefit.
  Unless
  explicitly varied, the parameters of the simulation are $N\!=\!100$,
  $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run
  for $T\!=\!5\times 10^6$ time steps for each parameter
  combination.}\label{fig:cooperation_rate_over_benefit}
\end{figure}

We also investigate the effect of the strength of selection.
Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process. For weak selection, \(\beta < 1\), we observe
that the two methods yield the same result, however, as \(\beta\) increases
there is variation in the evolving populations. In the case of expected payoffs
the resident populations become more and more cooperative as \(\beta\)
increases, whereas in the case of limited memory payoffs, the resident
populations become less cooperative.

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.65\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_selection_strenght.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_selection_strenght.pdf}
  \end{subfigure}%
  \begin{subfigure}{.35\textwidth}
    \vspace{.3cm}
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different selection strength values.} 
We vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
probabilities \(p, q\) for resident population over \(10^7\) time steps for each
\(\beta\) value. ({\bf B}) The cooperation rate within the resident population
(averaged over all games and over all time steps) over \(\beta\).
Unless explicitly varied, the parameters of the simulation are $N\!=\!100$,
$b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
$T\!=\!5\times 10^6$ time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

@@

In the previous subsection we gained insights into the effects of the updating
payoffs, and into how parameters such as the benefit and the strength of
selection can intensify them. We investigated these effects by using the
donation game. In order to broaden our understanding of the updating payoffs on
different forms of possible human interactions, we extend our approach to other
\(2 \times 2\) symmetric games. More specifically, we apply our analysis to the
four different classes of games given by Table~\ref{table:social_dilemmas}. We
compare the results of the evolutionary process described in
section~\ref{section:model} when the updating payoffs are based on the expected
payoffs, and on the last round payoffs for all the four possible classes of
games. The results of the simulations are given in
Figures~\ref{fig:expected_two_by_two} and~\ref{fig:last_round_two_by_two}
respectively.

\begin{table}[!htbp]
  \begin{center}
    \resizebox{.45\textwidth}{!}{
    \renewcommand{\arraystretch}{1.5}
      \begin{tabular}{ccc}
  \toprule
  & \textbf{social dilemmas} & \textbf{preference ordering} \\
  \midrule
(iii) & harmony  & \(R > T > S > P\) \\
(i) & stag hunt  & \(R > T > P > S\) \\
(iv) & prisoner dilemma  & \(T > R > P > S\) \\
(ii) & snowdrift  & \(T > R > S > P\) \\
  \bottomrule
      \end{tabular}}
  \end{center}
\caption{\textbf{Social dilemmas and payoffs' constrains}. The various social
dilemmas we explore in this work. Results for cases (i) - (iv) are presented
in section~\ref{section:2_by_2_games} and results for case (v) are presented
in section~\ref{section:donation}.}
\label{table:social_dilemmas}
\end{table}

In Figure~\ref{fig:expected_two_by_two} each sub figure represents a run of
evolutionary process for a different set of values the payoffs of an individual,
\(\mathcal{U}\). Without loss of generality we set \(R=1\) and
\(P=0\)~\cite{Martinez2012, Roca2009}, and we vary the values of the temptation
\(T\) (across the \(x\)-axis) and of the risk of cooperating \(S\) (across the
\(y\)-axis). There are four classes of games as presented in
Table~\ref{table:social_dilemmas}. Starting at the upper left corner and proceeding
clockwise correspond to the harmony game, the snowdrift, the prisoner's dilemma
and the stag hunt.

The harmony game represents a situation without tensions where it is in the best
interest of both players to cooperate. This is confirmed by our simulations
Figure~\ref{fig:expected_two_by_two}. In the several experiments of the harmony
game the resident population applies a strategy for which \(p \approx 1\) and
\(q \approx 1\).

The snowdrift game describes a situation similar to the prisoner's dilemma where
cooperation results in a benefit to the opposing player but entailing a cost to
the cooperator. Compared to the prisoner's dilemma, individuals obtain immediate
direct benefits from the cooperative acts. The story of the snowdrift game
usually goes as follow: two drivers are trapped on either side of a snowdrift
and have the options of staying in the car or removing the snowdrift. Letting
the opponent do all the work is the best option but if the other player stays in
the car it is better to shove. In the end cooperation leads to the cooperated
drive to get home, even though they had to endure the work, thus  \(S > P\). In
the snowdrift game more cooperation emerges compared to the prisoner's dilemma.
Defection is never a resident strategy and for the shame values of temptation
the overall \(q-values\) are higher. It can be seen that the \(p-values\) are
lower as individuals have free room to defect after receiving a cooperation.

The last class of games corresponds to the stag hunt game. In the stag hunt game
both players benefit for mutual cooperation and defection, however, \(R > P\) and
thus the resident strategies with the highest fitness are the cooperative ones.
The results are similar to the cases of the harmony game.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/expected_two_by_two_games.pdf}
  \caption{{\bf Evolutionary dynamics under expected payoffs for various social dilemmas.} 
  We have run several simulations of the evolutionary process described in
  section~\ref{section:model} for $T\!=\!10^7$ time steps. The graphs show how
  often the resident population chooses each combination ($p,q$) of conditional
  cooperation probabilities in the subsequent rounds. We vary the temptation
  payoff \(T \in \{-1, -0.6, -0.2,  0.2, 0.6, 1, 1.4, 1.8, 2.2, 2.6, 3\}\)
  across the \(x\) axis, and  \(S \in \{2, 1.6, 1.2, 0.8, 0.4, 0, -0.4, -0.8,
  -1.2, -1.6, -2\}\) across the \(y\) axis. Parameters: $N\!=\!100$, $c\!=\!1$,
  $\beta\!=\!1$, $\delta\!=\!0.999$.}
  \label{fig:expected_two_by_two}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/stochastic_two_by_two_games.pdf}
  \caption{{\bf Evolutionary dynamics under last round payoffs for various social dilemmas.} 
  We have run several simulations of the evolutionary process described in
  section~\ref{section:model} for $T\!=\!10^7$ time steps. The graphs show how
  often the resident population chooses each combination ($p,q$) of conditional
  cooperation probabilities in the subsequent rounds. We vary the temptation
  payoff \(T \in \{-1, -0.6, -0.2,  0.2, 0.6, 1, 1.4, 1.8, 2.2, 2.6, 3\}\)
  across the \(x\) axis, and  \(S \in \{2, 1.6, 1.2, 0.8, 0.4, 0, -0.4, -0.8,
  -1.2, -1.6, -2\}\) across the \(y\) axis. Parameters: $N\!=\!100$, $c\!=\!1$,
  $\beta\!=\!1$, $\delta\!=\!0.999$.}
  \label{fig:last_round_two_by_two}
\end{figure}

The results for the last round payoffs are fairly similar with the patterns
begging the same. It can been seen from Figure~\ref{fig:last_round_two_by_two}
that the results are move variation. We have discussed the expected behaviour
for each social dilemma. We now want to account the difference in the
cooperation rate given the two updating payoffs. The cooperation rates for each
of the case of the social dilemmas we have used are given in
Figure~\ref{fig:cooperation_rates_two_by_two_part_one}.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=.4\textwidth]{static/expected_two_by_two_games_cooperation.pdf}
  \includegraphics[width=.4\textwidth]{static/stochastic_two_by_two_games_cooperation.pdf}
  \caption{\textbf{Cooperation rates over various social dilemmas.} ({\bf A}) If
  players update based on their expected payoffs. ({\bf B}) If
  players update based on their last round payoffs.}\label{fig:cooperation_rates_two_by_two_part_one}
\end{figure}

>>>>

\begin{abstract}
Repeated games have been the centerpiece when studying the evolution of human
cooperation. Most of the existing models work with idealized assumptions.
Individuals are assumed to interact with a representative sample of the
population, and they remember all interactions they participate in. In real
life, however, individuals do not always remember all interactions they had with
everyone in their population. Instead, they rather recall the last interaction
they had.

In this work we explore the effect on the evolution of cooperation if
individuals only remember a minimum of social information. We present results
for a series of simulations were we vary the benefit of cooperation and the
strength of selection. We present results for the most commonly used classes of
social dilemmas, and we show that even though cooperation can evolve when
individuals only use a minimum of information, the evolving cooperation rates
are typically lower than in the classical scenario.
\end{abstract}

@@

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others,
increasing their payoff, at the expense of decreasing one’s own? In evolutionary
game theory individuals are not required to be rational, instead they adapt
strategies based on mutation and exploration. Strategies are more likely to
spread if they have a high fitness either because the individuals who adopt them
have more offspring, or because they are imitated more often~\cite{Wu2015}. The
fitness of a strategy is not constant but depends on the composition of the
population. Individuals interact based on their strategies with other members of
the population and the payoffs they yield are translated into fitness.

It is commonly assumed that individuals compute their fitness after interacting
with a representative sample of the population, and remembering all the
interactions they participated in~\cite{Nowak2006}. Thus, they imply that
individuals have a perfect memory. However, when modelling how individuals make
decisions in each turn they are assumed to have very limited memory. To be
precise, most of the works in the literature focus on naive subjects who can
only choose from a restricted set of strategies~\cite{Nowak1992tit}, or who do
not remember anything beyond the outcome of the very last round~\cite{Baek2016}.
Note that there are a few notable exceptions~\cite{Hauert1997, Stewart2016}.
The perfect memory assumption is not only unrealistic but it also creates
this curious inconsistency.

This has led us to question the robustness of our understanding of cooperation.
In this work we explore whether cooperation can evolve if individuals compute
their fitness based on a minimum of social information. Though we are not the
first to question the assumptions of estimating fitness~\cite{Roca2006}, we are
the first to explore the effect of payoff memory. %ToDo add more information

@@

The remainder of the paper is organized as follows. In
section~\ref{section:model} we describe the model. In
section~\ref{section:results} we present the results of the simulations, and
in section~\ref{section:conclusions} we outline the main conclusions.

@@

We consider a population of \(N\) players~\footnote{The terms ``player'' and
``individual'' are used interchangeably here.} where \(N\) is even and mutations
are sufficiently rare. Therefor, at any point in time there are at most two
different strategies present in the population; a \textit{resident} strategy and
a \textit{mutant} strategy. To describe how strategies spread we use a pairwise
comparison process~\cite{Traulsen2006}.
Each step of the evolutionary process consists of two
stages, a game stage and an updating stage.

In the game stage each individual is randomly matched with some other individual
in the population. They engage in a match where each subsequent turn
occurs with a fixed probability $\delta$. At each turn players can choose to
either cooperate (\(C\)) or to defect (\(D\)), and the payoffs of the turn
depend on both their decisions. If both players cooperate they receive the
reward payoff \(R\), whereas if both defect they receive the punishment payoff
\(P\). If one cooperates but the other defects, the defector receives the
temptation \(T\) payoff, whereas the cooperator receives the sucker's
payoff \(S\). We denote the feasible payoff of each turn as \(\mathcal{U} =
\{R, S, T, P\}\).
We assume that individuals use \textit{reactive strategies} to make
decisions in each turn. Reactive strategies are a set of memory-one strategies
that only take into account the previous action of the opponent. They can be
written explicitly as a vector \(\in \R_{3}\), more specifically, a reactive
strategy \(s\) is given by \(s=(y, p, q)\) where \(y\) is the probability that
the strategy opens with a cooperation and \(p, q\) are the probabilities that
the strategy cooperates given that the opponent cooperated and defected
equivalently.

In the updating stage, two players are randomly drawn from the population, a
`learner' and a `role model'. Given that the learner's payoff $u_L\!\in\!
\mathcal{U}$ and that the role model's payoff $u_{RL}\!\in\! \mathcal{U}$.
The learner adopts the role model's strategy based on the Fermi
distribution function,

@@

where $\beta\!\ge\!0$ is the intensity of selection. The updating payoffs of the
learner and the role model are conventionally correspond to their expected
payoffs. The expected payoff of a strategy is the mean payoff of the strategy of
engaging in repeated games with all other members of the population. The expected payoffs
assume perfect memory. In this work we define a new a set of updating
payoffs. We reefer to these as the limited memory payoffs, and we compare them
to the classical scenario.

The evolutionary step is repeated until either the mutant strategy goes
extinct, or until it fixes in the population. If the mutant fixes in the
population then the mutant strategy becomes the new resident strategy. After
either outcome we introduce a new mutant strategy uniformly chosen from all
reactive strategies at random, and we set the number of mutants to $1$. This
process of mutation and fixation/extinction is then iterated many times.
More details on our methodology are found in Appendix~\ref{appendix:methods}.

In order to account for the effect of the updating payoffs approaches, we
explore the cooperation rate within the resident population over multiple
generations. In the next section we present a series of simulation results.
To account for the various types of social behaviour we also present
results on multiple social dilemmas.

@@

In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population and we compare
this to the classical scenario of the expected payoffs. We assume that each pair
of players interacts in a donation game. The donation game is a special case of
the prisoner's dilemma. Each player can choose to cooperate by providing a
benefit \(b\) to the other player at their cost \(c\), with \(0 < c < b\).
Thus, the feasible payoffs in each round are \(T=b, R=b-c, S=-c, P=0\).

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulations results
for the described process of section~\ref{section:model}.
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$. The discount factor~$\delta$
is comparably high, thus we do not report the opening move \(y\) as it
is a transient effect. The left panels correspond to the
standard scenario considered in the literature, it considers players who use
expected payoffs to update their strategies. The right panel shows the scenario
considered herein, in which players update their strategies based on their last
round’s payoff. The top panels assume a benefit \(b\) of 3 whereas the bottom
assume a benefit of 10.

The figure suggests that when updating is based on expected payoffs players
tend to be more generous and more cooperative. The $q$-values of the resident
strategies are on average higher in the case of the expected payoffs. The
players will occasionally forgive a defection more often if their fitness
depends on interacting with every member of the population. On the other hand,
when social interactions are limited they are less forgiving. The average
cooperation rate for each simulation is calculated as the average
cooperation rate within the resident population. In the case of the expected
payoffs, regardless the value of benefit, the average cooperation rate is
strictly higher than that of the last round payoffs. The difference based on the
two methods is statistically significant, and in the case of $b=10$ the
average cooperation of resident strategies drops from 97\% to 57\%.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.70\textwidth]{static/expected_and_stochastic_for_donation_game.pdf}\vspace{-3cm}
    \includegraphics[width=.70\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and last round with one interaction payoffs.} 
    We have run two simulations of the evolutionary process described in
    section~\ref{section:model} for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since simulations
    are run for a relatively high continuation probability of $\delta\!=\!0.999$, we
    do not report the players' initial cooperation probability $y$. The graphs show
    how often the resident population chooses each combination ($p,q$) of
    conditional cooperation probabilities in the subsequent rounds. ({\bf A}) If
    players update based on their expected payoffs, the resident population
    typically applies a strategy for which $p\!\approx\!1$ and
    $q\!\le\!1\!-\!c/b\!=\!0.9$. ({\bf B})
    When players update their strategies based on their realized payoffs in the last
    round, there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$ consistent
    with stable cooperation is somewhat smaller compared to the expected-payoff
    setting, $q\!<\!0.5$. The cooperation rate within the resident population
    (averaged over all games and over all time steps) is close to 100\%. Parameters:
    $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

We further explore the effect of benefit in
Figure~\ref{fig:cooperation_rate_over_benefit}. The figure suggests that
expected payoffs always yield a higher cooperation rate. In the case of expected
payoffs we observe that the cooperation rate increases as the value of the
benefit gets higher. In comparison for the limited memory payoffs, the
cooperation rate remains unchanged at approximately 50\% once \(b=5\).

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=.8\textwidth]{static/expected_for_beta.pdf}
    \includegraphics[width=.8\textwidth]{static/stochastic_for_beta.pdf}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=.8\textwidth]{static/cooperation_rate_over_b.pdf}
  \end{subfigure}
  \caption{{\bf The evolution of cooperation for different benefit values.} 
  We vary the benefit of defection $b$. In all cases, expected payoffs appear to
  overestimate the average cooperation rate the population achieves. ({\bf A})
  the probabilities \(p, q\) for resident population over \(10^7\) time steps
  for each benefit value. ({\bf B}) The cooperation rate within the resident population
  (averaged over all games and over all time steps) over the benefit.
  Unless
  explicitly varied, the parameters of the simulation are $N\!=\!100$,
  $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run
  for $T\!=\!5\times 10^7$ time steps for each parameter
  combination.}\label{fig:cooperation_rate_over_benefit}
\end{figure}

We also investigate the effect of the strength of selection $\beta$.
Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process. For weak selection, \(\beta < 1\), we observe
that the two methods yield similar results, however, as \(\beta\) increases
there is variation in the evolving populations. In the case of expected payoffs
the resident populations become more cooperative as \(\beta\)
increases, whereas in the case of limited memory payoffs, the resident
populations become more defective.

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.65\textwidth}
    \centering
    \includegraphics[width=.8\textwidth]{static/expected_for_selection_strenght.pdf}
    \includegraphics[width=.8\textwidth]{static/stochastic_for_selection_strenght.pdf}
  \end{subfigure}%
  \begin{subfigure}{.35\textwidth}
    \vspace{.3cm}
    \includegraphics[width=.8\textwidth]{static/cooperation_rate_over_betas.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different selection strength values.}
We vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
probabilities \(p, q\) for resident population over \(10^7\) time steps for each
\(\beta\) value. ({\bf B}) The cooperation rate within the resident population
(averaged over all games and over all time steps) over \(\beta\).
Unless explicitly varied, the parameters of the simulation are $N\!=\!100$,
$b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
$T\!=\!5\times 10^7$ time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

@@

In the previous section we gained insights into the effects of the updating
payoffs, and into how parameters such as the benefit and the strength of
selection can intensify them. We investigated these effects by using the
donation game. In order to broaden our understanding of the updating payoffs on
different forms of possible human interactions we extend our approach to other
\(2 \times 2\) symmetric games. More specifically, we apply our analysis to four
different classes of games as given by Table~\ref{table:social_dilemmas}.

\begin{table}[!htbp]
  \begin{center}
    \resizebox{.4\textwidth}{!}{
    \renewcommand{\arraystretch}{1.5}
      \begin{tabular}{ccc}
  \toprule
  & \textbf{social dilemmas} & \textbf{preference ordering} \\
  \midrule
(i) & harmony  & \(R > T > S > P\) \\
(ii) & stag hunt  & \(R > T > P > S\) \\
(iii) & prisoner dilemma  & \(T > R > P > S\) \\
(iv) & snowdrift  & \(T > R > S > P\) \\
  \bottomrule
      \end{tabular}}
  \end{center}
\caption{\textbf{Social dilemmas and preference ordering}. The four classes of
social dilemmas we explore in this work.}
\label{table:social_dilemmas}
\end{table}

We compare the results of the evolutionary process described in
section~\ref{section:model} when the updating payoffs are based on the expected
payoffs, and on the last round payoffs for all the four possible classes of
games. The results of the simulations are given in
Figures~\ref{fig:expected_two_by_two} and~\ref{fig:last_round_two_by_two}
respectively.

In Figure~\ref{fig:expected_two_by_two} each sub figure represents a run of
evolutionary process for a different set of values for \(\mathcal{U}\). Without
loss of generality we set \(R=1\) and \(P=0\)~\cite{Martinez2012, Roca2009}, and
we vary the values of the temptation payoff \(T\) (across the \(x\)-axis) and of
the sucker's payoff \(S\) (across the \(y\)-axis). Starting at the upper left
corner and proceeding clockwise the quadrants correspond to the harmony, the
snowdrift, the prisoner's dilemma and the stag hunt games.

The harmony game represents are social situation without conflict. It is in the
best interest for both players to cooperate as \(R > T\) and \(S > P\).
Figure~\ref{fig:expected_two_by_two} confirms this. We observe that for most
harmony game runs the resident population overwhelmingly applies a strategy for
which \(p\) and \(p\) are \(\approx 1\). In the case of \(T = R = 1\) we observe
that the resident strategies become slightly less generous. Against a cooperating
strategy defecting and cooperation yield the same result. The lower \(q-\)values
could indicate that the resident strategy is less generous to defectors to avoid
being invaded.

The snowdrift game describes a situation similar to that of the prisoner's
dilemma. Cooperation results in a benefit to the opposing player but entailing a
cost to the cooperator. However, in the snowdrift game individuals obtain
immediate direct benefits from the cooperative acts which leads to \(S > P\). In
the snowdrift game more cooperation emerges compared to the prisoner's dilemma.
Defection is never a resident strategy and for the shame values of temptation
the overall \(q-values\) are higher. It can be seen that the \(p-values\) are
lower as individuals have free room to defect after receiving a cooperation.

The last class of games we present are for the stag hunt game. In the stag hunt
both players benefit for mutual cooperation and defection, however, \(R > P\)
and thus the resident strategies with the highest fitness are the cooperative
ones. The results for the stag hunt class are similar to those of the harmony
game.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/expected_two_by_two_games.pdf}
  \caption{{\bf Evolutionary dynamics under expected payoffs for various social dilemmas.} 
  We have run several simulations of the evolutionary process described in
  section~\ref{section:model} for $T\!=\!10^7$ time steps. The graphs show how
  often the resident population chooses each combination ($p,q$) of conditional
  cooperation probabilities in the subsequent rounds. We vary the temptation
  payoff \(T \in \{0.2, 0.6, 1, 1.4, 1.8, 2.2\}\)
  across the \(x\) axis, and  \(S \in \{1.2, 0.8, 0.4, 0, -0.4, -0.8\}\)
  across the \(y\) axis. Unless explicitly varied, the parameters of the simulation
  are $N\!=\!100$, $\beta\!=\!10$, $\delta\!=\!0.99$.}
  \label{fig:expected_two_by_two}
\end{figure}

The conclusions regarding the evolved behaviour for the various classes of
games remain the same when the last round payoffs are considered,
Figure~\ref{fig:last_round_two_by_two}. However, there is variation in these
populations. This variation as an effect affects the resident within cooperative
rates.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\textwidth]{static/stochastic_two_by_two_games.pdf}
  \caption{{\bf Evolutionary dynamics under last round payoffs for various social dilemmas.} 
  We have run several simulations of the evolutionary process described in
  section~\ref{section:model} for $T\!=\!10^7$ time steps. The graphs show how
  often the resident population chooses each combination ($p,q$) of conditional
  cooperation probabilities in the subsequent rounds. We vary the temptation
  payoff \(T \in \{0.2, 0.6, 1, 1.4, 1.8, 2.2\}\)
  across the \(x\) axis, and \(S \in \{1.2, 0.8, 0.4, 0, -0.4, -0.8\}\)
  across the \(y\) axis. Unless explicitly varied, the parameters of the simulation
  are $N\!=\!100$, $\beta\!=\!10$, $\delta\!=\!0.99$.}
  \label{fig:last_round_two_by_two}
\end{figure}

The cooperation rates for each of the case of the social dilemmas we have used
are given in Figure~\ref{fig:cooperation_rates_two_by_two_part_one}. In the case
of the harmony game the differences are very small, even in the case of \(T=1\)
where the biggest difference occurs it still remain less than 10\%. In the case
of the stag hunt game the difference increases in the case of \(T=1\). The
biggest variation in the results are for the class of the prisoner's dilemma.
For values of \(S < -0.4\) the are significantly different, in the case of the
last round payoffs cooperation almost never is a valid option. Overall the last
round payoffs are less cooperative, supporting the results of the previous
section. There are only two cases which that is not true and both cases are in
the snowdrift class. For the values of \(T>1.8\) and \(S=0.4, 0.8\) cooperation
is estimated higher. In the last round payoffs a Tit For Tat like behaviour is
more robust that in the case of the expected payoffs.
